---
title: "Three Logicians Walk Into a Bar"
output:
  html_document:
    df_print: paged
---

```{r echo=F}
library(lattice)
library(latticeExtra)
```
# Intro

### The Joke

*Three logicians walk into a bar. The bartender asks, "Do you all want a beer?"*

-   *Ada: "I don't know"*
-   *Bertrand: "I don't know"*
-   *Claude: "Yes!"*

### Explanation

How does this work? Each logician really does want a beer, but doesn't know what the others want. If the bartender's question is taken literally ("do **all** of you want a beer), no logician knows the answer - at first. (On other hand, if any of the logicians didn't want a beer, he or she could answer,"no".) After Ada and Bertrand give their "I don't know" answers (thus - they do want a beer), Claude can give a definite answer.

### Generalizations

#### Forms of the Joke

There are other forms of this joke, depending on the bartender's question. An alternative, equally funny, but less jolly form of the joke is:

*Three logicians walk into a bar. The bartender asks, "do any of you want a beer?"*

-   *A: "I don't know"*
-   *B: "I don't know"*
-   *C: "No!"*

In this case, no one wanted a beer, but again only Claude was able to answer the bartender.

Here's a third form of the joke:

*Three logicians walk into a bar. The bartender asks, "do an odd number of you want a beer?"*

-   *A: "I don't know"*
-   *B: "I don't know"*
-   *C: "I don't know!"*

So funny!

#### Counting the Jokes
The bartender could also ask some less bartenderly questions, like "does A want a beer, and does B not want a beer, and does C want a beer?" How many bartender questions are there? Well... 256.

To see why, consider this plot of the first question, "do you all want a beer"? The plot shows the final answer, if all logicians' desires were known.
```{r echo=F}
dat <- data.frame(
  expand.grid(x=c("A yes B no", "A yes B yes", "A no B yes", "A no B no"), y=c("C no", "C yes")),
  value=c(0, 0, 0, 0, 0, 1, 0, 0)+1,
  source=c("no", "no", "no", "no", "no", "yes", "no", "no")
)
lp <- levelplot(value ~ x+y, data=dat, xlab=NULL, ylab=NULL, colorkey=NULL,
  col.regions=colorRampPalette(colors=c("red", "green"))(2))
xp <- xyplot(y ~ x, data=dat,
  panel = function(y, x, ...) {
    ltext(x = x, y = y, labels = dat$source, cex = 1, font = 2,
      fontfamily = "HersheySans")
  }
  )

p <- lp + xp
print(p)
```

And here's "do any of you want a beer?":
```{r echo=F}
dat <- data.frame(
  expand.grid(x=c("A yes B no", "A yes B yes", "A no B yes", "A no B no"), y=c("C no", "C yes")),
  value=c(1, 1, 1, 0, 1, 1, 1, 1)+1,
  source=c("yes", "yes", "yes", "no", "yes", "yes", "yes", "yes")
)
lp <- levelplot(value ~ x+y, data=dat, xlab=NULL, ylab=NULL, colorkey=NULL,
  col.regions=colorRampPalette(colors=c("red", "green"))(2))
xp <- xyplot(y ~ x, data=dat,
  panel = function(y, x, ...) {
    ltext(x = x, y = y, labels = dat$source, cex = 1, font = 2,
      fontfamily = "HersheySans")
  }
  )

p <- lp + xp
print(p)
```

Last example, "do an odd number of you want a beer?":
```{r echo=F}
dat <- data.frame(
  expand.grid(x=c("A yes B no", "A yes B yes", "A no B yes", "A no B no"), y=c("C no", "C yes")),
  value=c(1, 0, 1, 0, 0, 1, 0, 1)+1,
  source=c("yes", "no", "yes", "no", "no", "yes", "no", "yes")
)
lp <- levelplot(value ~ x+y, data=dat, xlab=NULL, ylab=NULL, colorkey=NULL,
  col.regions=colorRampPalette(colors=c("red", "green"))(2))
xp <- xyplot(y ~ x, data=dat,
  panel = function(y, x, ...) {
    ltext(x = x, y = y, labels = dat$source, cex = 1, font = 2,
      fontfamily = "HersheySans")
  }
  )

p <- lp + xp
print(p)
```

Each arrangement of the plot has 8 squares that can be "yes" or "no". So the number of questions is  $2^{8} = 256$.

#### Effect of the Need for Beer

Our three logicians may or may not want a beer; 3 logicians and 2 options means there are ${2}^{3}$ = `r 2**3` ways the joke could unfold. But! Some of these outcomes are identical. For example, if it's "do you all want a beer?" and A doesn't want a beer, the answers are "no", "no", and "no", no matter what B and C want. If A wants a beer but B doesn't, the answers are "I don't know", "no", and "no", even if C wants a beer.

Here's a plot showing all 8 possible joke outcomes.
```{r echo=F}
# joke result, "and" form, per logician desire
# the "backwards" row/col numbers make
# the data match the input format
m <- matrix(nrow=3, ncol=8, data=c(
   0, -1, -1,  # 100
   0,  0, -1,  # 110
  -1, -1, -1,  # 010
  -1, -1, -1,  # 000
   0, -1, -1,  # 101
   0,  0,  1,  # 111
  -1, -1, -1,  # 011
  -1, -1, -1   # 001
))

levelplot(
  m,
  col.regions=colorRampPalette(colors=c("red", "yellow", "green")),
  xlab="answers",
  ylab="want beer?",
  #colorkey=NULL,
  colorkey=list(at=c(-1, -.33, .33, 1), labels=list(at=c(-0.67, 0, 0.67), labels=c("No", "I don't know", "Yes"))),
  scales=list(
    x=list(at=1:3, labels=c("A", "B", "C")),
    y=list(at=1:8, labels=c(
      "A yes B no C no", "A yes B yes C no", "A no B yes C no", "A no B no C no",
      "A yes B no C yes", "A yes B yes C yes", "A no B yes C yes", "A no B no C yes"
    ))
  )
)
```

#### Interlude: Notation

For shorthand, let's say the "Do you wall want a beer" joke is $j_{AND}$; "Do any of you want a beer?" is $j_{OR}$, and "Do an odd number of you want a beer, $j_{XOR}$. The logicians' beer/no-beer choice is 1 for beer, 0 for no-beer, and there are three of them, so we can use a triplet of 0/1 to show what they want. We can encode the answer as"I don't know" $\mapsto$ $X$, "no" $\mapsto$ $0$, "yes" $\mapsto$ $1$. Putting it all together, the first form the joke above is $$j_{AND}(1, 1, 1) =  \{X, X, 1\}$$

The other 7 outcomes of this form of the joke are: $$j_{AND}(0, 0, 0) =  \{0, 0, 0\}$$ $$j_{AND}(0, 0, 1) =  \{0, 0, 0\}$$ $$j_{AND}(0, 1, 0) =  \{0, 0, 0\}$$ $$j_{AND}(0, 1, 1) =  \{0, 0, 0\}$$ $$j_{AND}(1, 0, 0) =  \{X, 0, 0\}$$ $$j_{AND}(1, 0, 1) =  \{X, 0, 0\}$$ $$j_{AND}(1, 1, 0) =  \{X, X, 0\}$$ The number of different ways $j_{AND}$ can be answered, $N_{AND}$, is 4: $\{0, 0, 0\}, \{X, 0, 0\}, \{X, X, 0\}, \{X, X, 1\}$.
